---
title: "0607-Lecture"
author: "Amber Potter"
date: "6/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

Exploring the Lahman dataset

Shortcut: command shift r - adds section label which allows for collapsible comments in R script file

## Packages

```{r}
library(tidyverse)
library(Lahman)
```


## Initial Examination of Data

```{r}
Batting <- as_tibble(Batting)
dim(Batting)
class(Batting)

head(Batting, 10)
tail(Batting, 10)

colnames(Batting)
names(Batting)
```

## EDA

```{r}
summary(Batting$yearID) # summarizing quantitative column
table(Batting$lgID) # summarizing categorical column

any(is.na(Batting$lgID)) # are any of vector elements true

mlb_batting <- Batting %>%
     filter(lgID == "AL" | lgID == "NL")
     # alternate filter: filter(lgID %in% c("AL", "NL"))

sel_batting <- Batting %>%
     dplyr::select(yearID, lgID, G, AB, R, H, HR, BB, SO)

arrange(Batting, desc(HR))

Batting %>%
     summarize(max(stint), min(stint))

new_batting <- Batting %>%
     mutate(batting_avg = H / AB)

# piping
# compute batting average
Batting %>%
     filter(lgID == "AL" | lgID == "NL", AB > 300) %>%
     mutate(batting_avg = H / AB) %>%
     arrange(desc(batting_avg)) %>%
     dplyr::select(playerID, yearID, batting_avg) %>%
     head(n = 5) # slice to access specific rows

Batting %>%
     filter(lgID == "AL" | lgID == "NL", AB > 300) %>%
     mutate(batting_avg = H / AB) %>%
     group_by(yearID) %>%
     summarize(hr = sum(HR), so = sum(SO), b = sum(BB), .groups = "drop") %>% 
     # use .groups = "drop" to get rid of previous groups - will use often
     # same as ungroup(), but ungroup is more outdated
     arrange(desc(hr)) %>%
     head(5)

year_batting_summary <- Batting %>%
     filter(lgID %in% c("AL", "NL")) %>%
     group_by(yearID) %>%
     summarize(total_hits = sum(H, na.rm = TRUE),
               total_hrs = sum(HR, na.rm = TRUE),
               total_ks = sum(SO, na.rm = TRUE),
               total_walks = sum(BB, na.rm = TRUE),
               total_abs = sum(AB,na.rm = TRUE), .groups = "drop") %>%
     mutate(batting_avg = total_hits / total_abs)


year_batting_summary %>%
     dplyr::select(yearID, batting_avg) %>%
     rename(Year = yearID, `Batting AVG` = batting_avg) %>%
     arrange(desc(`Batting AVG`)) %>%
     slice(1:3, n()) %>% # n() is used within tidy functions, otherwise use nrows
     gt() %>%
     tab_header()


```

## Notes: 

- R attempt at object oriented programming uses @ for accessing objects
- reference variables with spaces included with back ticks surrounding the name
- n() is used within tidy functions, otherwise use nrows



