---
title: "Lecture-0608"
author: "Amber Potter"
date: "6/9/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## New dataset - 2021 MVP Shohei Ohtani's batted balls

Created dataset of batted balls by the American League MVP Shohei Ohtani in 2021 season using [`baseballr`](http://billpetti.github.io/baseballr/):

```{r load-ohtani, warning = FALSE, message = FALSE}
library(tidyverse)
ohtani_batted_balls <- 
  read_csv("http://www.stat.cmu.edu/cmsac/sure/2022/materials/data/sports/xy_examples/ohtani_2021_batted_balls.csv")
head(ohtani_batted_balls)
```

--
- each row / observation is a batted ball from Ohtani's 2021 season

--
- __Categorical__ / qualitative variables: `pitch_type`, `batted_ball_type`, `outcome`

--
- __Continuous__ / quantitative variables: `hit_x`, `hit_y`, `exit_velocity`, `launch_angle`

---

## Visualizing 1D categorical data

How can we summarize `batted_ball_type` and other categorical variables?

--
- We make a __bar chart__ with [`geom_bar()`](https://ggplot2.tidyverse.org/reference/geom_bar.html)

```{r batted-ball-bar}
ohtani_batted_balls %>%
  ggplot(aes(x = batted_ball_type)) +
  geom_bar() + #<<
  theme_bw()
```

- Only map `batted_ball_type` to the x-axis

- Counts of each type are displayed on y-axis...


## Remember statistical summaries!

From [Chapter 3 of `R` for Data Science](https://r4ds.had.co.nz/data-visualisation.html)

---

## What does a bar chart show?

__Marginal distribution__: probability that categorical variable X (e.g., `batted_ball_type`) takes each particular value x (e.g. `fly_ball`); how a single column of our dataset vary marginally within complete dataset.
--
  _So how do we display the individual probabilities?_
--

```{r batted-ball-bar-prop}
ohtani_batted_balls %>%
  ggplot(aes(x = batted_ball_type)) +
     # probability/proportion on y axis
  geom_bar(aes(y = after_stat(count) / sum(after_stat(count)))) + #have to do this in geom_bar not ggplot
  theme_bw()
```

- [`after_stat()`](https://ggplot2.tidyverse.org/reference/aes_eval.html) indicates the aesthetic mapping is performed after the statistical transformation

- Use `after_stat(count)` to access the `stat_count()` called by `geom_bar()`

- __We can code this in a more clear way__

---

## Compute and display the proportions directly


```{r batted-ball-bar-prop2}
ohtani_batted_balls %>%
  group_by(batted_ball_type) %>% #<<
  summarize(count = n()) %>% #<<
  ungroup() %>% #<<
  mutate(total = sum(count), #<<
         prop = count / total) %>% #<<
  ggplot(aes(x = batted_ball_type)) +
  geom_bar(aes(y = prop), #<<
           stat = "identity") + # for using previously calculated y value
  theme_bw()
```

- Category counts give info about sample size, but this could be labeled in the chart

- Proportions $=$ the __probability mass function__ (PMF) for __discrete__ variables 
  - e.g. $P$ (`batted_ball_type` $=$ `fly_ball`)

---

## Population versus sample...

We have the __population__ of Ohtani's batted balls in the 2021 season
--
$\Rightarrow$ __we know the true probabilities__:

- $P$ (`batted_ball_type` $=$ `fly_ball`) 
- $P$ (`batted_ball_type` $=$ `ground_ball`)
- $P$ (`batted_ball_type` $=$ `line_drive`)
- $P$ (`batted_ball_type` $=$ `popup`)


--
_What if we pretend this is a sample from all hypothetical Ohtani 2021 seasons_?


--
__Empirical distribution__: We __estimate__ the __true marginal__ distribution with __observed (sample) data__

Estimate $P$ (`batted_ball_type` = $C_j$) with $\hat{p}_j$ for each category $C_j$ (e.g. $\hat{p}_{\texttt{fly_ball}}$)


--
Compute __standard error__ for each $\hat{p}_j$:

--

$$
SE(\hat{p}_j) = \sqrt{\frac{\hat{p}_j ( 1 - \hat{p}_j)}{n}}
$$

For large $n$ $\Rightarrow$ $\approx$ 95% __confidence interval (CI)__: $\hat{p}_j +/- 2 \cdot SE(\hat{p}_j)$

---

## Add confidence intervals to bar chart


```{r bar-intervals}
ohtani_batted_balls %>%
  group_by(batted_ball_type) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(total = sum(count),
         prop = count / total,
         se = sqrt(prop * (1 - prop) / total),  #<<
         lower = prop - 2 * se,  #<<
         upper = prop + 2 * se) %>% #<<
  ggplot(aes(x = batted_ball_type)) +
  geom_bar(aes(y = prop),
           stat = "identity") + 
  geom_errorbar(aes(ymin = lower, #<<
                    ymax = upper), #<<
                color = "red") + #<<
  theme_bw()
```

__Be careful about your interpration of CIs...__

_You should remember to label your charts!_


## Fun with factors using [`forcats`](https://forcats.tidyverse.org/)


```{r bar-factors}
ohtani_batted_balls %>%
  group_by(batted_ball_type) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(total = sum(count),
         prop = count / total,
         se = sqrt(prop * (1 - prop) / total),  
         lower = prop - 2 * se, 
         upper = prop + 2 * se,
         batted_ball_type = #<<
           fct_reorder(batted_ball_type, #<< reorder variables
                       desc(prop))) %>% #<< default ascending
  ggplot(aes(x = batted_ball_type)) +
  geom_bar(aes(y = prop),
           stat = "identity") + 
  geom_errorbar(aes(ymin = lower, 
                    ymax = upper),
                color = "red") + 
  theme_bw()
```

---

## Did you say pie chart?

[](https://cdn.vox-cdn.com/uploads/chorus_asset/file/3500550/0003VL-9150.0.jpeg)

--
__This is the only pie chart I will show you all summer__

---

## Describing 1D continuous data

How can we summarize `exit_velocity` and other continuous variables?

--
- __Center__: mean, median, number and location of modes

- __Spread__: range (max - min), quantiles, variance (standard deviation), etc.

- __Shape__: skew vs symmetry, outliers, heavy vs light tails, etc.

- Compute basic summary statistics

```{r exit-velo-summary}
summary(ohtani_batted_balls$exit_velocity)


sd(ohtani_batted_balls$exit_velocity) # does not work because of NAs
sd(ohtani_batted_balls$exit_velocity, na.rm = TRUE)
```


---

## Box plots visualize summary statistics


- We make a __box plot__ with [`geom_boxplot()`](https://ggplot2.tidyverse.org/reference/geom_boxplot.html)


### Visualize exit velocity

```{r boxplot}
ohtani_batted_balls %>%
  ggplot(aes(y = exit_velocity)) + #<<
  geom_boxplot(aes(x = "")) + #<< for when we make a single boxplot to avoid numerical x axis values
  theme_bw() +
  coord_flip() #<< good for single boxplot
```

- __Pros__:
  - Displays outliers, percentiles, spread, skew
  - Useful for side-by-side comparison (tomorrow)

- __Cons__:
  - Does not display the full distribution shape!
  - Does not display modes
  
boxplots have very limited use because they dont show the shape of the data beyond a simple intuition about the direction of a tail

histograms are the most useful to show the shape and distribution of our data
  
---

## Histograms display 1D continuous distributions


- We make __histograms__ with [`geom_histogram()`](https://ggplot2.tidyverse.org/reference/geom_histogram.html)

```{r histogram}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity)) + #<<
  geom_histogram() + #<<
  theme_bw() 
```


$$
\text{# total obs.} = \sum_{j=1}^k \text{# obs. in bin }j
$$

- __Pros__:
  - Displays full shape of distribution
  - Easy to interpret

- __Cons__:
  - Have to choose number of bins and bin locations (will revisit later)
     - this is an estimate, we have to guess what the true distribution looks like when we choose bins. different numbers of bins makes the distribution look different
---

## Display the data points directly with beeswarm plots



- We make a __beeswarm plot__ using the  [`ggbeeswarm` package](https://github.com/eclarke/ggbeeswarm)


```{r beeswarm}
library(ggbeeswarm)
ohtani_batted_balls %>%
  ggplot(aes(y = exit_velocity)) + 
  geom_beeswarm(aes(x = ""), #<<
                cex = 3) + #<<
  theme_bw() +
  coord_flip()
```

- __Pros__:
  - Displays each data point
  - Easy to view full shape of distribution
  - nice for small datasets

- __Cons__:
  - Can be overbearing with large datasets -> can potentially get around this by plotting a random sample of your points
  - Which algorithm for arranging points?
  - points on the bottom offer redundant information

_What does `cex = 3` do?_

---

## Smooth summary with violin plots


- We make __violin plots__ with [`geom_violin()`](https://ggplot2.tidyverse.org/reference/geom_violin.html)

```{r violin}
ohtani_batted_balls %>%
  ggplot(aes(y = exit_velocity)) + 
  geom_violin(aes(x = "")) + #<<
  theme_bw() +
  coord_flip()
```

- __Pros__:
  - Displays full shape of distribution
  - Can easily layer...
  

Beeswarm and violin plots are normally not used by themselves. 

Add a boxlplot on top of a violin plot. Use a smaller width for the boxpots so it does not cover the violin plot.
---

## Smooth summary with violin plots + box plots

- We make __violin plots__ with [`geom_violin()`](https://ggplot2.tidyverse.org/reference/geom_violin.html)

```{r violin-box, eval = FALSE}
ohtani_batted_balls %>%
  ggplot(aes(y = exit_velocity,
             x = "")) + 
  geom_violin() + 
  geom_boxplot(width = .2) + #<<
  theme_bw() +
  coord_flip()
```

- __Pros__:
  - Displays full shape of distribution
  - Can easily layer... with box plots on top
  
- __Cons__:
  - Summary of data via __density estimate__ 
  - Mirror image is duplicate information


---

### What do visualizations of continuous distributions display?

__Probability that continuous variable X takes a particular value is 0__ 

e.g. $P$ (`exit_velocity` $= 100$) $= 0$, _why_?


--
Instead we use the __probability density function (PDF)__ to provide a __relative likelihood__ 

- Density estimation is the focus of lecture next Monday


--
For continuous variables we can use the __cumulative distribution function (CDF)__,

$$
F(x) = P(X \leq x)
$$

--
For $n$ observations we can easily compute the __Empirical CDF (ECDF)__:


$$\hat{F}_n(x)  = \frac{\text{# obs. with variable} \leq x}{n} = \frac{1}{n} \sum_{i=1}^{n}1(x_i \leq x)$$

- where $1()$ is the indicator function, i.e. `ifelse(x_i <= x, 1, 0)`,(Note: indicator function 1 does not require special font)

---



## Display full distribution with ECDF plot

- We make __ECDF plots__ with [`stat_ecdf()`](https://ggplot2.tidyverse.org/reference/stat_ecdf.html)

```{r ecdf}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity)) + 
  stat_ecdf() + #<<
  theme_bw()
```

- __Pros__:
  - ECDF displays all information in data (except for order)
  - As $n \rightarrow \infty$, our ECDF $\hat{F}_n(x)$ converges to the true CDF $F(x)$
  - Easy to interpret...
  
- __Cons__:
  - ... and yet it's not as popular! (probably because it is cumulative)
  
---

## Rug plots display raw data


- We make a __rug plot__ with [`geom_rug()`](https://ggplot2.tidyverse.org/reference/geom_rug.html)


```{r rugplot}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity)) + #<<
  geom_rug(alpha = 0.7) + #<<
  theme_bw()
```

- __Pros__:
  - Displays raw data points
  - Useful supplement for summaries and 2D plots...

- __Cons__:
  - Can be overbearing for larger datasets

---

## Rug plots supplement other displays



```{r rugplot-hist}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity)) + #<<
  geom_rug(alpha = 0.7) + #<<
  geom_histogram() + #<<
  theme_bw()
```


```{r rugplot-ecdf}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity)) + #<<
  geom_rug(alpha = 0.7) + #<<
  stat_ecdf() + #<<
  theme_bw()
```

**histogram plots the derivative of the ecdf plot**

---

## Scatterplots for 2D continuous data

Joint distributions

- We make a __scatterplot__ with [`geom_point()`](https://ggplot2.tidyverse.org/reference/geom_point.html)


```{r scatter}
ohtani_batted_balls %>%
  ggplot(aes(x = exit_velocity, #<<
             y = launch_angle)) + #<<
  geom_point(alpha = 0.4) + #<<
  geom_rug(alpha = 0.4) + #<<
  theme_bw()
```

_Easy to supplement with rug plots_ (dont need to include rug plots on every plot)

__Look at the plot__: what question would you want to ask, assuming you know something about baseball?

_To be continued..._
  