---
title: "eda-lab-0613"
author: "Amber Potter"
date: "6/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

```{r}
library(tidyverse)
library(wehoop)
library(flexclust)
library(ggrepel)
library(protoclust)
```

## Purpose: Exploring WNBA Data

```{r}
wnba_pbp_data <- load_wnba_pbp(2022)

# Get the shots and clean this data a bit:
wnba_shots_data <- wnba_pbp_data %>%
  filter(shooting_play)

# Remove unnecessary columns:
wnba_shots_data <- wnba_shots_data %>%
  dplyr::select(-shooting_play, -id, -participants_2_athlete_id,
                -type_abbreviation, -season, -season_type, 
                -home_team_spread, -game_spread, -home_favorite)

# Save this file:
write_csv(wnba_shots_data, 
          "data/wnba_shots_2022.csv")
```

### Initial Ideas

- home vs away - bar graph position = "fill"

- points over time - facet by team, plot home and away average scores over time

- points by team

- first half vs second half - bar graph, position = "fill"

- shot type - regroup/recombine shot type names - standard bar graph

- scatter plot - time on x axis, each shot is a point, colored


-----------

### Ideas

- shot type frequency or accuracy percentage - bar graph (position = "fill")

- points over time - facet by team, plot home and away average scores over time - line graph (smooth or linear???)

- x and y coordin.- scatter/hex/heatmap


## EDA Hypotheses

### Home Team Advantage

Does the home team have a scoring advantage?

  - Look at slopes of point totals throughout each game.

```{r}
wnba_shots_data <- wnba_shots_data %>%
     mutate(total_seconds = ((((period_number - 1) * 10) * 60) + 
                               ((10 - as.numeric(clock_minutes)) * 60) + 
                               (60 - as.numeric(clock_seconds)))/60,
            away_team = paste0(away_team_name, " ", away_team_mascot),
            home_team = paste0(home_team_name, " ", home_team_mascot),
            shooting_team = case_when(team_id == home_team_id ~ home_team,
                                     team_id == away_team_id ~ away_team),
            home_away = case_when(team_id == home_team_id ~ "Home",
                                     team_id == away_team_id ~ "Away"),
            team_score = case_when(team_id == home_team_id ~ home_score,
                                     team_id == away_team_id ~ away_score))


wnba_shots_data %>%
     filter(scoring_play) %>%
     ggplot(aes(x = total_seconds, y = team_score, color = home_away)) +
     geom_point(alpha = .1) +#+
     geom_smooth(se = FALSE, alpha = .7, method = "lm") +
     ggthemes::scale_color_tableau() +
     facet_wrap(~shooting_team) +
     theme_minimal() +
     labs(x = "Minutes",
          y = "Team Points",
          color = "Home or Away",
          title = "Being the home team is not as much of an advantage as one might think")
```



### Shot Accuracy

What are the most accurate shots?

  - Look at percentage of baskets and misses, also look at shot frequency

```{r}
wnba_shots_data <- wnba_shots_data %>%
     mutate(shot_type = case_when(grepl("Layup", type_text, fixed = TRUE) ~ "Layup",
                                  grepl("Jump Shot", type_text, fixed = TRUE) ~ "Jump Shot",
                                  grepl("Free Throw", type_text, fixed = TRUE) ~ "Free Throw",
                                  grepl("Tip Shot", type_text, fixed = TRUE) ~ "Tip Shot",
                                  grepl("Hook", type_text, fixed = TRUE) ~ "Hook Shot",
                                  grepl("Dunk", type_text, fixed = TRUE) ~ "Dunk"))

wnba_shots_data_simp %>%
     group_by(shot_type) %>%
     filter(n() > 100) %>%
     summarize(shot_type,
               prop = mean(scoring_play), 
               scoring_play,
               .groups = "drop") %>%
     ggplot(aes(x = reorder(shot_type, -prop), y = prop, fill = scoring_play)) +
     geom_bar(stat = "identity", position = "fill") +
     labs(x = "Shot Type",
          y = "Proportion",
          fill = "Did they score?") +
  theme_minimal() +
  theme(legend.position = "bottom")

wnba_shots_data_simp %>%
     group_by(shot_type, scoring_play) %>%
     summarize(count = n(), .groups = "drop") %>%
     pivot_wider(names_from = scoring_play, 
                 names_prefix = "scored_",
                 values_from = count,
                 values_fill = 0) %>%
    mutate(total_freq = scored_TRUE + scored_FALSE) %>%
    knitr::kable(col.names = c("Shot Type", "Frequency Scored", "Frequency Missed", "Total Fequency")) %>%
    kable_styling(bootstrap_options = "striped", full_width = TRUE, position = "center", font_size = 16) # looks better in presentation
```

### Shot locations

Where are most shots taken from?

```{r}
wnba_shots_data_location <- wnba_shots_data %>%
     filter(coordinate_x >= 0,
            coordinate_y >= 0)


#All shots (made and not made) -----------------------------------------------
wnba_shots_data_location %>%
  ggplot(aes(x = coordinate_x, y = coordinate_y)) +
  geom_hex(binwidth = c(2,2)) + 
  scale_fill_gradient(low = "darkblue", high = "darkorange") +
  theme_minimal() +
          ylim(min = -1, max = 70) +
     facet_wrap(~scoring_play) +
  theme(legend.position = "bottom") + 
  labs(x = "X Coordinate (ft)", y = "Y Coordinate (ft)",
       fill = "Frequency") + 
  coord_equal()
```



## Clustering

### Clusters of different shot types based on average distance and average points

***K-means++:***

```{r}
# Pts per shot vs shot distance for different shot types 
avg_pts_distance <- wnba_shots_data %>%
  mutate(scorediff = case_when(home_away == "Home" ~ home_score - away_score,
                               home_away == "Away" ~ away_score - home_score),
         shot_distance = sqrt((coordinate_x - 25)^2 + (coordinate_y)^2)) %>%
  filter(coordinate_x >= 0,
            coordinate_y >= 0) %>%
  group_by(type_text) %>%
  summarize(n = n(), 
            avgpts = mean(score_value), 
            avgdist = mean(shot_distance)) %>%
  filter(n >= 50) %>%
  mutate(std_avgdist = as.numeric(scale(avgdist)),
         std_avgpts = as.numeric(scale(avgpts))) %>%
  arrange(desc(avgpts)) 
  
kmeanspp <-
  kcca(dplyr::select(avg_pts_distance,
                     std_avgdist, std_avgpts),
       k = 3, control = list(initcent = "kmeanspp"))

avg_pts_distance %>%
  mutate(shot_clusters = as.factor(kmeanspp@cluster)) %>%
  ggplot(aes(x = avgdist, y = avgpts, color = shot_clusters)) +
  geom_point() +
  geom_text_repel(label = avg_pts_distance$type_text, size = 3.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  ggthemes::scale_color_colorblind() +
  labs(x = "Average Distance from Basket", 
       y = "Average Points Return",
       color = "Shot Clusters")
```


***H-clust (centroid):***

```{r}
type_dist <- dist(dplyr::select(avg_pts_distance, std_avgdist, std_avgpts))

# Centroid linkage
type_hclust <- hclust(type_dist, method = "centroid")

avg_pts_distance %>%
  mutate(type_clusters = as.factor(cutree(type_hclust, k = 3))) %>%
  ggplot(aes(x = avgdist, y = avgpts, color = type_clusters)) +
  geom_point() +
  geom_text_repel(label = avg_pts_distance$type_text, size = 3.5) +
  labs(x = "Average Distance from Basket", y = "Average Points Return") +
  theme_minimal() +
  theme(legend.position = "none") +
  ggthemes::scale_color_colorblind()
```

***Minimax***

```{r}
shot_type_dist <- dist(dplyr::select(avg_pts_distance,
                                  std_avgdist, std_avgpts))

shot_minimax <- protoclust(shot_type_dist)

minimax_shot_clusters <- protocut(shot_minimax, k = 3)

avg_pts_distance %>%
  mutate(shot_clusters =
           as.factor(minimax_shot_clusters$cl)) %>%
  ggplot(aes(x = avgdist, y = avgpts, color = shot_clusters)) + 
  geom_point(alpha = .6) + 
  geom_text_repel(label = avg_pts_distance$type_text, size = 3.5) +
  labs(x = "Average Distance from Basket", y = "Average Points Return") +
  theme_minimal() +
  theme(legend.position = "none") +
  ggthemes::scale_color_colorblind()
```

